Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2023-02-25 ms/ms03_results.tex
ENH

For the genome-wide association studies (GWAS), I used data from the IEU OpenGWAS database, that provides GWAS summary statistics
from different batches such as the EBI database of complete GWAS summary data, manually collected GWAS and UK Biobank.
%
GWAS traits were selected from the IEU OpenGWAS database based on four criteria \cite{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 417 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% TODO: Add to supplementary table?
% select count(distinct gwas_id) from tophits
% 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

The colocalization pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples in the somatic chromosomes 1-22 \cite{2019.Posthuma.Watanabe,2021.Li.Mu}.
%
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
Variants in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000 were
excluded from the colocalization pipeline, because of their complex linkage disequilibrium structure.
%
The histogram shows three groups of GWAS (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}).
%
There were 100\% of leading variants explained with eQTLs for between 5\% and 12\% of the GWAS depending on whether
we set PP.H4.SNP at 0.25, 0.5 or 0.75 (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc} and suppl. fig. S1).
This parameter gives the posterior probability of a SNP in a window to be causal.
%
There were a middle group of maximally 5\% GWAS with maximum percentage of explained leading SNPs between 10\% and 50\%
depending on the PP.H4.SNP parameter 0.25-0.75
Around 25\% of GWAS had no tophits colocalized with eQTLs for a parameter PP.SNP.H4=0.5.
%
Finally, between 17\% and 27\% GWAS had 0\% of leading variants explained by eQTLs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc} and suppl. fig. S1).

@@

These variants annotated with GWAS association and gene expression associations in particular biosamples are useful functional informations.
%
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0) as sel
%138301
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
%5343
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

@@

% TODO CONTINUE
These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC

% TODO add rs2522051
This variant rs2522051 is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of the breast, circulatory system diseases, hypertension, and respiratory system diseases
(Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1
%
This variant rs2522051 is active in 23 classes of tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4, and RAD50.

>>>>

For the genome-wide association studies (GWAS), I used data from the IEU OpenGWAS database, that provides GWAS summary statistics
from different batches such as the EBI database of complete GWAS summary data, manually collected GWAS and UK Biobank.
%
GWAS traits were selected from the IEU OpenGWAS database based on four criteria \cite{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 417 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS.
% select count(distinct gwas_id) from tophits
% 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

The colocalization pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples in the somatic chromosomes \cite{2019.Posthuma.Watanabe,2021.Li.Mu}.
%
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
Variants in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000 were
excluded from the colocalization pipeline, because of their complex linkage disequilibrium structure.
%
The histogram shows three groups of GWAS depending on the percentage of explained leading SNPs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}a).
%
For around 25\% of GWAS, there are not colocalizations between the leading SNPs and eQTLs.
%
On the other side, for around 7.5\% of GWAS, 100\% of the leading SNPs colocalize with eQTLs.
%
The remaining GWAS show between 10\% and 50\% colocalized leading SNPs with eQTLs.

The number of colocalization depends on the PP.H4.SNP parameter that gives the probability of a SNP in the colocalized locus to be the causal SNP.
%
We have changed this parameter between 25\% and 75\%, which results in 5\% and 12\% of the GWAS
showing colocalization with 100\% of their leading variants.
%
In this range, between 16\% and 25\% of the GWAS do not show any colocalization.
%
The remaining GWAS show between 5\% and 10\% of their leading variants colocalized with eQTLs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}a and Supplementary figure S1).

@@

These variants annotated with GWAS associations and eQTLs are useful functional information.
%
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0) as sel
%138301
%select count(*) from (select DISTINCT chrom,pos,alt FROM coloc
%where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
%5343
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

I provide a UCSC track hub with these GWAS colocalizations.
%
In Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}b-d, I show three screenshots of this UCSC track hub for three different biosamples, namely B-cells, naive monocytes and frontal cortex biosamples in the chromosome 5 around the gene IL4.
%
This screenshot shows that depending on the biosample, the regulated genes are very different.
%
In the

@@

These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC

% TODO add rs2522051
This variant rs2522051 is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of the breast, circulatory system diseases, hypertension, and respiratory system diseases
(Table \ref{tab:pleitropic_variants}a and Supplementary table 4).
%
The variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1
%
This variant rs2522051 is active in 23 classes of tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4, and RAD50.
